--- 
title: "The very basics of programming"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  pdf_document: default
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(fig.path = './assets/')
```

One of the very nice features of R when one comes from other statistical
software like SAS or Stata is that it is very easy to write customized code and
develop your own functions. That is a very good thing because the manipulation,
analysis, and visualization of data is considerably easier when one can write 
small functions. In addition, because R is open-source, you can inspect what
every function does, so it is useful to get a sense of the most basic elements 
of programming to be able to take advantage of that feature.

### Defining functions

Think about functions as a way of packing operations in one reusable expression.
For instance, consider a trivial example in which we want to calculate a mean.
Instead of doing `sum` over `length` in every situation, we could just pack
those operations together and give them a name:

```{r}
my_mean <- function(x) sum(x)/length(x)
my_mean(c(0, 1))
```

To define a function, we use the keyword `function` followed by parentheses, 
(optionally) the arguments that the function takes, and the expression that the
function runs. We then assign this function to a name, in this case, `my_mean`.

See how typing `my_mean` allows you to see what you have just defined.

```{r}
my_mean
```

The way we defined the function is perfectly valid but we could also be a bit
more explicit by enclosing the statement in parenthesis and ensuring that it is
returned.

```{r}
my_mean <- function(x) {
  return(sum(x)/length(x))
}
```

Note the use of braces to enclose the contents of the function. Braces are
optional when the function body consists of a single line but are necessary when
the functional body contains multiple lines.

### Conditionals

As you can imagine, conditionals execute an action depending on the value 
(`TRUE` or `FALSE`) of another statement. In R, they use the follow syntax:

```{r eval=FALSE}
if (x > 0) {
  print("It's positive") 
} else if (x == 0) {
  print("It's neither positive nor negative")
} else {
  print("It's negative")
}
```

which we may wrap up in a function
```{r}
my_sign <- function(x) {
  if (x > 0) {
    out <- "It's positive!"
  } else if (x == 0) {
    out <- "It's neither positive nor negative!"
  } else {
    out <- "It's negative!"
  }
  return(out)
}
```

and use it to illustrate what happens when you are not careful:[^1]
```{r}
my_sign("A cow")
```

How to deal with the previous issue? By issuing an error:

```{r}
my_sign <- function(x) {

  if (!is.numeric(x)) {
    stop("Input is not a number")
  }
  
  if (x > 0) {
    out <- "It's positive!"
  } else if (x == 0) {
    out <- "It's neither positive nor negative!"
  } else {
    out <- "It's negative!"
  }
  return(out)
}
```

Two things two notice here. First, that `!` is the negation operator (`TRUE ==
!FALSE`). Second, that `stop` interrupts the evaluation and produces an error,
so the function never reaches the next conditional.

### Vectorized operations

One common need is to apply an operation to each individual element of a vector
or  list. As in most programming languages, this can be done in R through looping
constructs. However, the use of loops in R is considered by some to be bad
practice, and loopos can significantly slow down the
execution of one's code in some cases, particularly when updated a data frame
within a loop. This is not always true, such as when adding elements to a 
allocated list, but even in
cases where loops do not increase run time, one should be hesitant to use loops
since they are not consistent with R's status as a (somewhat) functional 
vector-based language. Another reason to avoid loops is that they can be 
difficult to translate into a parallelized version.

What do we do then? The alternative in R is vectorized operations. 
There are a number of ways to do this. First, let's define a function that 
calculates a factorial of a number.


```{r eval = FALSE}
myfactorial <- function(x) {

  if(x == 0) {
    return(1)
  } else {
    return(x*myfactorial(x-1))
  }
  
}
```

Note that this function calls itself. This is known as recursion. In practice,
some checks on the parameter should probably be added to assure that it's a 
nonnegative integer, and there are surely faster, safer implementations, but
let's not worry about that for now. Let's instead consider what we would need
to do to apply the function to each element of a vector. We cannot simply
pass in the vector because the function expects a scalar, i.e., a vector of
length 1. There are a few ways around this.

One option is to use the `apply` family of functions. These are functions that
can apply a function to individual elements of a vector or list, or apply a
function to each row or column of a matrix.  In this case, we would use 
`sapply`.


```{r eval = FALSE}
results <- sapply(1:5, myfactorial)
```

The first parameter to `sapply` is the vector, and second parameter is the
function which we want to apply the vectors elements. `sapply` will return
a vector containing the resulting factorials. Note that if the return values
are not all of the same type, then `sapply` will instead return a list. If you
want a predictable return type, one can use `lapply` instead, which works the 
same way but always returns a list.

There are other functions in the apply family, such as `apply`, which applies
a function to every row or column of a matrix, but they are used in a similar
manner. Further, some R users prefer a group of functions defined in the `plyr` 
and `dplyr` packages which perform apply-like operations. These packages are
briefly discussed in the next section.

A second option is to create a new function which handles all of these `apply`
operations automatically. This can be done through the `Vectorize` function. We
would define a new vectorized version like this:



```{r eval = FALSE}
myfactorial_vectorized <- Vectorize(myfactorial)
```

Alternatively, we could have done this in our original function definition.

```{r eval = FALSE}
myfactorial <- Vectorize(function(x) {

  if(x == 0) {
    return(1)
  } else {
    return(x*myfactorial(x-1))
  }
  
})
```

Using our new "vectorized" function is simple:

```{r eval = FALSE}
results <- myfactorial_vectorized(1:5)
```

Please note that `Vectorize` has one important parameter that we did not
mention: `vectorize.args`. This should be set to a vector of strings which
contain the variable names we wish to "vectorize" over. For example, suppose
we have the following function:


```{r eval = FALSE}
weird_sum <- function(x, y, z) {

  x + y + sum(z)
  
}

```

Now assume we want to vectorize the function over `x` and `y` but not `z`, i.e.,
we want to add the first element of `x`, the first element of `y`, and the sum of
`z`, and then the second element of `x`, the second element of `y`, and the sum
of `z`, and so on. We would do it like this:


```{r eval = FALSE}
weird_sum <- Vectorize(function(x, y, z) {

  x + y + sum(z)
  
}, vectorize.args = c("x", "y"))

```

*Actually, `weird_sum` will work as intended without the use of `Vectorize` but 
only because the `+` operator is vectorized. For the sake of illustration, let's
ignore this pesky detail.*

We did not need to specify `vectorize.args` in our `myfactorial` example because
the default value is to vectorize over all arguments, which is what we wanted
in that case.

In some cases, both options are unnecessary because there
are already functions to accomplish our task. For example, the built-in 
`factorial` function in R is already vectorized. Similarly, there are a
number of built-in functions for calculating sums and means over rows or 
columns of a matrix, e.g., `rowSums`, `colMeans`. The use of these kind of 
functions is simpler and often results in faster programs than the above, so
one should use explicit vectorization only as needed.